---
title: "Must prediction in sdm be made on same data using for training model?"
author: "Wyclife Agumba Oluoch"
date: "<center>`r Sys.time()`</center>"
bibliography: 
  - bib/packages.bib
nocite: '@*'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---


```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
packages <- c("base",
              'knitr',
              'rmarkdown',
              'sdm',
              'raster')
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == FALSE)){
  install.packages(packages[!installed_packages])
}
lapply(packages, library, character.only = TRUE) |> invisible()
```

```{r write_bib, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::write_bib(c(
  .packages(), packages
), 'bib/packages.bib')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Background

I am writing this short article to clear air on whether a change in one or all of raster layers used in building a model will prevent prediction using the trained model.

If a model is trained using say stacked raster layers, say four in total, if values of one raster is changed, the same model will still perform prediction.

Most important is to ensure that the names of the rasters used in training the model are maintained even if values are changed.

Here, I will use data that comes with `sdm` package to enable ease of reproducibility.

# Loading data

```{r}
file <- system.file("external/species.shp", package = "sdm")
species <- shapefile(file)
path <- system.file("external", package = "sdm")
lst <- list.files(path = path, pattern = 'asc$', full.names = T)
preds <- stack(lst)
```

# Creating sdmData object

```{r}
sdm_data <- sdmData(formula = Occurrence ~., train = species, predictors = preds)
```

# Creating the model

```{r}
sdm_model <- sdm(Occurrence ~., data = sdm_data, methods = c('rf', 'svm'), replication = 'boot', n = 3)
```

# Makiing prediction on the original data

```{r}
prediction_original <- predict(sdm_model, preds)
plot(prediction_original[[2]])
```

# Changing one of the rasters in preds stack

```{r}
elevation <- preds[['elevation']]
elevation_doubled <- elevation * 2 
```

# Feeding the new raster back to preds stack

```{r}
preds[['elevation']] <- elevation_doubled
preds_2 <- preds # Preds with one raster values doubled
```

# Making prediction with one changed raster in the stack
```{r}
prediction_changed <- predict(sdm_model, preds_2) 
plot(prediction_changed[[2]])
```

# Ploting the two outputs

```{r}
par(mfrow = c(2,1))
plot(prediction_original[[2]])
plot(prediction_changed[[2]])
```

































